TIMES Ireland Model: Residential Sector

### 2021-06-01 manuscript.tex
Residential ENergy Modelling in Progress

\title{TIMES Ireland Model: Residential Sector}
% The changing TIMES of Irish Homes
\author[inst1,inst2]{Jason Mc Guire\footnote{Contact: j.mcguire@ucc.ie}}

@@

A €15/ $tCO_2$ carbon tax was introduced in 2009 to reduce investment in fossil fuels, and the increasing carbon tax rate has resulted in tax receipts growing every year from 2010 to 2018 \cite{Oireachtas2019AnOffice}. Ireland's current carbon tax of €33.50/ $tCO_2$ will rise to €100/ $tCO_2$ in 2030. The Renewable Energy Feed-in Tariff (REFIT) schemes were designed to incentivise renewable electricity investment to achieve legally binding renewable targets under 2009/28/EC. REFIT operates by guaranteeing a minimum price for electricity over a 15 year period. The Renewable Electricity Support Scheme (RESS) replaced REFIT in 2020. RESS supports to renewable electricity projects to help Ireland achieve 2030 climate targets, especially the 70\% renewable electricity target. 

@@

\subsection{Residential Modelling}
\label{Intro:ESOM}
%Convert the ESOM section in the intro to a “residential modelling section” which I would use to “describe” how residential sector has been model with TIMES and other models (I put describe in “” because I would just mention the approaches and cite specific studies), as well as mention their advantages/disadvantages;- Olex Email
Energy System Optimisation Models (ESOMs) can be used to optimise the selection of a large number of dynamic variables to satisfy demand, while accounting for complex sectoral interconnections within each time period.
\par Some of the main ESOM frameworks include EnergyPLAN \cite{stergaard2015ReviewingSimulations}, MESSAGE \cite{Messner1995WorkingE}, TIMES \cite{Loulou2016DocumentationI}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOYS \cite{Howells2011OSeMOSYS:Development.}.

While ESOM can cover all the energy-related sectors, it can also be used to explore the residential sector in detail.

Some studies focusing on the residential sector in TIMES include, Li who studied the preference for heating technology choice in UK \cite{Li2018IncorporatingModel}, Petrovic examined a new approached to represent residential heat pumps in Denmark \cite{Petrovic2016ResidentialSystem} and Aberg analysed the affect higher thermal efficiency in buildings has on co-production of electricity and district heating in Sweden \cite{Aberg2011OptimisationBuildings}. There has also been some studies residential sector in Ireland, using previous versions of Irish TIMES and now out-dated climate targets.

@@

\subsection{TIM Overview}
\label{Overview}
NOT DONE 
The Integrated MARKAL-EFOM System (TIMES) modelling framework was utilised to develop an Irish model, a description and selection process of the framework are outlined in the following subsection. 
\par TIMES has been developed and maintained by Energy Technology Systems Analysis Program (ETSAP) of the International Energy Agency (IEA). TIMES is a bottom-up optimisation model generator for energy-environment systems analysis at various levels of spatial, temporal and sectoral resolution.
\par TIMES uses The GAMS source code to optimise a least cost solution with perfect foresight. The objective function may be thought of as the discounted sum of net annual costs (i.e. costs minus revenues),
\begin{equation} 
\begin{split}
NPV = {\sum_{r=1}^{R}  \sum_{y\in YEARS}(1 + d_{r,y})^{REFYR-y}}
\cdot {ANNCOST(r,y)} 
\end{split}
%\[ \sum_{n=1}^{\infty} 2^{-n} = 1 \]
\label{Eq:ObjectiveFunction}
\end{equation}
where: $NPV$ is the net present value of the total cost for all regions (the TIMES objective function);
$ANNCOST(r,y)$ is the total annual cost in region r and year y; $d_{r,y}$ is the general discount rate;
$REFYR$ is the reference year for discounting;
YEARS is the set of years for which there are costs, including all years in the horizon, plus past years (before the initial period) if costs have been defined for past investments, plus a number of years after end of the horizon EOH where some investment and dismantling costs are still being incurred, as well as the Salvage Value; and
R is the set of regions in the area of study.
$ANNCOST$ is quite complex and is postponed until section 6.2 of PART II
Due to the data driven nature of TIMES (see section 2.5), all TIMES constraints are activated and defined by specifying some attributes. Attributes are attached to processes, to commodities, to flows, or to special variables that have been created to define new TIMES features. Indeed, TIMES

@@

The TIM approach.....data and calibration
%a newly formed group with members from different universities, industries and semi-state agencies within Ireland.  Following this, other datasets were then used to develop the full housing stock, based on the correlation of year of construction and building energy ratings.
\label{Results}
\subsubsection{Scenarios}
Behavioral Constraints here 
\subsection{Residential}
\label{LitR:RES}
ENERGY CONSUMPTION CHARACTERISTICS of Houses - 
 The BER database, which represents half of the dwelling stock, highlights the affect year of construction has on expected energy consumption, as shown in Fig.\ref{fig:BERAge}. The BERs labels are categorised based on expected energy consumption, with A1 representing the lowest expected energy consumption ($\leq 74 kWh/m^2/year$) and G representing the highest expected energy consumption ($ >450 kWh/m^2/year$). The BERs are developed from the Energy Performance of Buildings Directive 2002/91/ EC, which helps to better understand the housing stock. BERs are required for new houses, rented houses, sold houses and renovated houses.

@@

Theoretically G-rated BERs consume more energy and consequently have higher energy bills to maintain a comfortable internal temperature, but in reality G-rated dwellings have low internal temperatures, which is a reason for Ireland's high excessive winter mortality \cite{Clinch2000HousingMortality}. Fuel Poverty which which is defined as when a \textit{``household spends 10\% or more of income to achieve adequate warmth"} is a major contributor to the high excessive winter deaths. An Element Energy report claimed 28\% of Irish households were in fuel poverty in 2015  \cite{ElementEnergy2015Bottom-upIreland}. Fuel Allowance and the Household Benefits provides support to low income households to alleviate fuel poverty. These payments have been increased over time with a further increase made in Budget 2020, coinciding with the increase in the carbon tax \cite{,EuropeanUnion2020EUObservatory}. Kerr's study finds an explicit link between fuel poverty and social welfare policy in Ireland \cite{,EuropeanUnion2020EUObservatory,Kerr2019PoliticsFrance} and O’Meara highlights the health issues associated with Fuel Poverty \cite{OMeara2016AIreland}. Ironically, it can be argued this places low income households deeply dependent on fossil fuels for health.
\par BERs do not take account of actual usage based on occupancy behaviour or occupancy density, nor does it take account of the size of the dwelling. The BER labels do therefore leave a gap between actual consumption and expected consumption, the literature review explains how this problem was tackled.
\subsubsection{Technology}
Capacity, Efficiency, Constraints, Uptake, adapation

\subsubsection{Costs}
\label{LitR:Costs}
\subsubsection{Savings}
\label{LitR:Savings}
\subsubsection{Demands}
\label{LitR:Demands}
Macroeconomic 
% textit{To accommodate an expanding population, the number of dwellings has also increased to currently stand at 1.7 million households, although this fgure remains signifcantly short of projected housing need, withannual average demand estimated to be up to 36,000 units for the next 30 years (IBEC 2018). Yet despite the number of new homes built to ever increasing energy performance standards, the Irish housing stock is among the poorest in Europe in terms of energy effciency (Goggins et al. 2016). Current trends also show that households are getting bigger, with an average 15\% increase in foor area across all homes between 2000 and 2016 (SEAI 2018) \cite{FahyEnergyPractice}}
Check main TIM paper 
Historical population estimates and future projections are obtained from the Central Statistics Office (CSO), Ireland \citep{CentralStatisticsOffice2020}. We use the M2F1 scenario since it represents medium growth in population and is in line with population projections used in other national sources \citep{Yakut2020}.

@@

%Space Heating demand - BER space heating is only within certain months and assumes 21C heating..
\subsubsection{Constraints}
\label{LitR:Constraints}

@@

\label{Results}
Electricity Supply, Technology and Fuel Mix, Retrofitting, Cost
\subsection{Trajectories}
\label{LitR:Traj}
%% The Appendices part is started with the command \appendix;
%% appendix sections are then done as normal sections
\section{Discussion}
\label{Discussion}
Soft-linking to LEAP and PLEXOS
\subsection{Strengths}
\label{Dis:Strengths}
\subsection{Limitations}
\label{Dis:Limit}
Consumers’ final investment decision can be
represented by “willingness-to-pay” curves – for
instance, around 10\% of owner-occupiers are willing
to invest in energy efficiency measures for a payback
time of 4 years, falling to 0\% for a payback time of
6 years for some consumer cohorts
\section{Conclusions}
\label{Conclusion}
This paper reports results on ambitious mitigation target in the period to 2050 for the Irish energy system. The analysis have been performed using the Irish TIMES model
\textit{We stress the importance of differentiating between technical con-
straints and behavioural constraints in optimisation models. Including both conflates what is considered desirable under the economic welfare paradigm (i.e. cost-minimising decisions) with “realistic decisions”, complicating the interpretation of the results. Optimisation models are by design well suited to study cost-optimal behaviour, and running scenarios just with technical constraints facilitates the identification of areas where market barriers are more entrenched, and interventions to accelerate the uptake of low-carbon practices may be needed.} \cite{Astudillo2017CanModelb}
\section{Acknowledgements}
\label{sec:Acknowledgements}
%CAPACITY is funded under the Climate Action Modelling Group (CAMG) by DECC
% https://www.marei.ie/project/capacity/
CHIMERA is supported by a research grant from Science Foundation Ireland (SFI) and the National Natural Science Foundation of China (NSFC) under the SFI-NSFC Partnership Programme, grant no. 17/NSFC/5181 and
supported by MaREI, the SFI Research Centre for Energy, Climate, and Marine [Grant No: 12/RC/2302\_P2]

>>>>

\title{TIMES Ireland Model: Residential Sector}
% The changing TIMES of Irish Homes

\author[inst1,inst2]{Jason Mc Guire\footnote{Contact: j.mcguire@ucc.ie}}

@@

Ireland uses both the \textit{carrot} (subsidy) and \textit{stick} (tax) approach to direct private invesvestment towards low carbon or renewable energy. A €15/ $tCO_2$ carbon tax was introduced in 2009 to reduce investment in fossil fuels, and the increasing carbon tax rate has resulted in tax receipts growing every year from 2010 to 2018 \cite{Oireachtas2019AnOffice}. Ireland's current carbon tax of €33.50/ $tCO_2$ will rise to €100/ $tCO_2$ in 2030. The Renewable Energy Feed-in Tariff (REFIT) schemes were designed to incentivise renewable electricity investment to achieve legally binding renewable targets under 2009/28/EC. REFIT operates by guaranteeing a minimum price for electricity over a 15 year period. The Renewable Electricity Support Scheme (RESS) replaced REFIT in 2020. RESS supports to renewable electricity projects to help Ireland achieve 2030 climate targets, especially the 70\% renewable electricity target. 

@@

\subsection{Residential Modelling}
\label{Intro:ESOM}

Energy System Optimisation Models (ESOMs) can be used to explore least cost pathways, as they consider a large amount of variables to satisfy demand, while accounting for complex sectoral interconnections within each time period. Some of the main ESOM frameworks include EnergyPLAN \cite{stergaard2015ReviewingSimulations}, MESSAGE \cite{Messner1995WorkingE}, TIMES \cite{Loulou2016DocumentationI}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOSYS \cite{Howells2011OSeMOSYS:Development.}.

While ESOM cover all the energy-related sectors, they can also be used to specifically explore the residential sector in detail. For example, Li et al. \cite{Li2018IncorporatingModel} investigated the heating technology preference in UK, using UK TIMES Model (UKTM). As TIMES assumes perfect foresight and \textit{``in reality, the behaviours of consumers is not always economically rational"} \cite{Li2018IncorporatingModel}. Li set out to incorporate a Discrete Choice Model (DCM) into TIMES to better reflect how homeowners choose heating technologies. The DCM was developed from a nationwide survey with 442 homeowners completing the survey. The nationwide survey identified the most influential factors for determining homeowners' preferences for heating systems. The DCM then estimated the probability of a heating technology selection and cost was found not to have a statistically significant impact on homeowners' choices. While Li's researcher has provided some initial insights, future work can improve the robustness and comprehensiveness of the insights, through more surveys and analysis.

Leibowicz et al. developed an expanded version of the Open Source Energy Modeling System (OSeMOSYS) that integrates electricity generation and storage, residential building energy service demands, and end-use technologies for providing those services \cite{Leibowicz2018OptimalServices}.
Endogenous constraints on the market penetration rate of each technology and hourly service demand profiles are established and applied to the residential sector of Austin, Texas. Austin was chosen because of the high granularity data availability \cite{EnergyInc.}. Leibowicz et al. find, net present costs in the \textit{``Policy scenarios"} ( emissions must decline linearly toward 10\% of their current level by 2050) are higher than \textit{``No Policy scenarios"} (No $CO_2$ regulations will be imposed over the model timeframe), but the percentage increase in net present cost due to the policy diminishes with a more thermally efficient building stock. Leibowicz et al. also finds electrification to be the dominant strategy for cost-effectively decarbonizing urban residential dwellings. 
These insights show the strength of an ESOM, but the not all costs are taken into account in Leibowicz study, the capital cost associated with more thermally efficient buildings are not included.

Hedegard and Münster used....

Patankar et al. TEMOA south sudan

Petrovic xamined a new approached to represent residential heat pumps in Denmark \cite{Petrovic2016ResidentialSystem} and Aberg analysed the affect higher thermal efficiency in buildings has on co-production of electricity and district heating in Sweden \cite{Aberg2011OptimisationBuildings}. There has also been some studies residential sector in Ireland, using previous versions of Irish TIMES and now out-dated climate targets.

@@

\subsection{TIMES Energy System Optimisation Model}
\label{Overview}
NOT DONE 
The Integrated MARKAL-EFOM System (TIMES) modelling framework was utilised to develop an Irish model, a description and selection process of the framework are outlined in the following subsection. 
\par TIMES has been developed and maintained by Energy Technology Systems Analysis Program (ETSAP) of the International Energy Agency (IEA). TIMES is a bottom-up optimisation model generator for energy-environment systems analysis at various levels of spatial, temporal and sectoral resolution.
\par TIMES uses The GAMS source code to optimise a least cost solution with perfect foresight. The objective function may be thought of as the discounted sum of net annual costs (i.e. costs minus revenues),
% The market impacts of grid tariffs are analysed using the energy
system model Balmorel (Wiese et al. (2018)) [25]. Balmorel is an open source, bottom-up, linear partial equilibrium model. It opti- mizes social economic surplus of production, transmission, con- sumption, and investments in electricity and DH under the assumption ofperfect competition, with perfect foresight, and with respect to exogenous technical, physical, and regulatory constraints per geographically defined area. Generation technologies have specified fuel types, efficiencies, fixed costs, variable costs, and environmental characteristics. CHP
\begin{equation} 
\begin{split}
NPV = {\sum_{r=1}^{R}  \sum_{y\in YEARS}(1 + d_{r,y})^{REFYR-y}}
\cdot {ANNCOST(r,y)} 
\end{split}
%\[ \sum_{n=1}^{\infty} 2^{-n} = 1 \]
\label{Eq:ObjectiveFunction}
\end{equation}
where: $NPV$ is the net present value of the total cost for all regions (the TIMES objective function);
$ANNCOST(r,y)$ is the total annual cost in region r and year y; $d_{r,y}$ is the general discount rate;
$REFYR$ is the reference year for discounting;
YEARS is the set of years for which there are costs, including all years in the horizon, plus past years (before the initial period) if costs have been defined for past investments, plus a number of years after end of the horizon EOH where some investment and dismantling costs are still being incurred, as well as the Salvage Value; and
R is the set of regions in the area of study.
$ANNCOST$ is quite complex and is postponed until section 6.2 of PART II
Due to the data driven nature of TIMES (see section 2.5), all TIMES constraints are activated and defined by specifying some attributes. Attributes are attached to processes, to commodities, to flows, or to special variables that have been created to define new TIMES features. Indeed, TIMES

@@

The TIM approach.....data and calibration
%a newly formed group with members from different universities, industries and semi-state agencies within Ireland.  Following this, other datasets were then used to develop the full housing stock, based on the correlation of year of construction and building energy ratings.
\label{Results}
\subsubsection{Scenarios}
Behavioral Constraints here 
\subsection{Residential}
\label{LitR:RES}
RES FIGURE
ENERGY CONSUMPTION CHARACTERISTICS of Houses - 
 The BER database, which represents half of the dwelling stock, highlights the affect year of construction has on expected energy consumption, as shown in Fig.\ref{fig:BERAge}. The BERs labels are categorised based on expected energy consumption, with A1 representing the lowest expected energy consumption ($\leq 74 kWh/m^2/year$) and G representing the highest expected energy consumption ($ >450 kWh/m^2/year$). The BERs are developed from the Energy Performance of Buildings Directive 2002/91/ EC, which helps to better understand the housing stock. BERs are required for new houses, rented houses, sold houses and renovated houses.

@@

Theoretically G-rated BERs consume more energy and consequently have higher energy bills to maintain a comfortable internal temperature, but in reality G-rated dwellings have low internal temperatures, which is a reason for Ireland's high excessive winter mortality \cite{Clinch2000HousingMortality}. Fuel Poverty which which is defined as when a \textit{``household spends 10\% or more of income to achieve adequate warmth"} is a major contributor to the high excessive winter deaths. An Element Energy report claimed 28\% of Irish households were in fuel poverty in 2015  \cite{ElementEnergy2015Bottom-upIreland}. Fuel Allowance and the Household Benefits provides support to low income households to alleviate fuel poverty. These payments have been increased over time with a further increase made in Budget 2020, coinciding with the increase in the carbon tax \cite{,EuropeanUnion2020EUObservatory}. Kerr's study finds an explicit link between fuel poverty and social welfare policy in Ireland \cite{,EuropeanUnion2020EUObservatory,Kerr2019PoliticsFrance} and O’Meara highlights the health issues associated with Fuel Poverty \cite{OMeara2016AIreland}. Ironically, it can be argued this places low income households deeply dependent on fossil fuels for health.
\par BERs do not take account of actual usage based on occupancy behaviour or occupancy density, nor does it take account of the size of the dwelling. The BER labels do therefore leave a gap between actual consumption and expected consumption, the literature review explains how this problem was tackled.
\subsubsection{House Stock Data}
BER Database, CSO and Calculations and Filtration

\subsubsection{Costs}
\label{LitR:Costs}
\subsubsection{Savings}
\label{LitR:Savings}
\subsubsection{Energy Service Demands Projections}
\label{LitR:Demands}
Macroeconomic 
Time slices - 40 etc. 
% textit{To accommodate an expanding population, the number of dwellings has also increased to currently stand at 1.7 million households, although this fgure remains signifcantly short of projected housing need, withannual average demand estimated to be up to 36,000 units for the next 30 years (IBEC 2018). Yet despite the number of new homes built to ever increasing energy performance standards, the Irish housing stock is among the poorest in Europe in terms of energy effciency (Goggins et al. 2016). Current trends also show that households are getting bigger, with an average 15\% increase in foor area across all homes between 2000 and 2016 (SEAI 2018) \cite{FahyEnergyPractice}}
Check main TIM paper 
Historical population estimates and future projections are obtained from the Central Statistics Office (CSO), Ireland \citep{CentralStatisticsOffice2020}. We use the M2F1 scenario since it represents medium growth in population and is in line with population projections used in other national sources \citep{Yakut2020}.

@@

%Space Heating demand - BER space heating is only within certain months and assumes 21C heating..
\subsubsection{Space and Water Heating}
Capacity, Efficiency, Constraints, Uptake, adapation
\subsubsection{Lighting, Appliances and Other}
Capacity, Efficiency, Constraints, Uptake, adapation
\subsubsection{User Constraints}
\label{LitR:Constraints}

@@

\label{Results}
Electricity Supply, Technology and Fuel Mix, Retrofitting, Cost

%% The Appendices part is started with the command \appendix;
%% appendix sections are then done as normal sections
\section{Discussion}
\label{Discussion}
Soft-linking to LEAP and PLEXOS
%o	Discuss and interpret the results
%o	Focus on what added value the model brings
%o	Also include future development priorities here
%o	What can’t the model do? Where is it more suitable to use another model - Hannah Email

\subsection{Strengths}
\label{Dis:Strengths}
\subsection{Limitations}
\label{Dis:Limit}
Consumers’ final investment decision can be
represented by “willingness-to-pay” curves – for
instance, around 10\% of owner-occupiers are willing
to invest in energy efficiency measures for a payback
time of 4 years, falling to 0\% for a payback time of
6 years for some consumer cohorts
\section{Conclusions}
\label{Conclusion}
This paper reports results on ambitious mitigation target in the period to 2050 for the Irish energy system. The analysis have been performed using the Irish TIMES model
\textit{We stress the importance of differentiating between technical con-
straints and behavioural constraints in optimisation models. Including both conflates what is considered desirable under the economic welfare paradigm (i.e. cost-minimising decisions) with “realistic decisions”, complicating the interpretation of the results. Optimisation models are by design well suited to study cost-optimal behaviour, and running scenarios just with technical constraints facilitates the identification of areas where market barriers are more entrenched, and interventions to accelerate the uptake of low-carbon practices may be needed.} \cite{Astudillo2017CanModelb}
\section{Acknowledgements}
\label{sec:Acknowledgements}
%CAPACITY is funded under the Climate Action Modelling Group (CAMG) by DECC
% https://www.marei.ie/project/capacity/
CHIMERA is supported by a research grant from Science Foundation Ireland (SFI) and the National Natural Science Foundation of China (NSFC) under the SFI-NSFC Partnership Programme, grant no. 17/NSFC/5181 and
supported by MaREI, the SFI Research Centre for Energy, Climate, and Marine [Grant No: 12/RC/2302\_P2]
